---
title: "Broad and fine scale factors predicting microclimates on coral reefs"
author: "Jaelyn Bos"
output:
  pdf_document: default
  html_document: default
date: 'Last updated: May 03, 2021'
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
library(dplyr)
library(ggplot2)
library(ggfortify)
library(mgcv)
library(caret)
library(ranger)
source('C:/github/aeda_bos/scripts/ML_helpers.R')
```
## Introduction  
  
Despite their small geographic area, coral reefs play an outsize role in maintaining marine biodiversity.^1^ Unfortunately, these spectacular ecosystems are seriously threatened by rising ocean temperatures^2^.  
However, coral reefs are thermally heterogeneous on multiple scales, and corals vary significantly in their thermal tolerance^3^. Some corals thrive in shallow seas or intertidal pools with routine high temperatures well above the bleaching thresholds of corals in more temperate locations^4^. In fact, conspecific corals inhabiting different temperature regimes can have different thermal tolerances, as measured in bleaching thresholds and growth at high temperatures^5^. This can be true not only at the inter-reef scale, but also at the intra-reef scale: corals inhabiting hot microclimates within reefs can have higher bleaching resistance than corals inhabiting more temperate areas of the same reef^6^.    
  Intriguingly, these effects sometimes persist intergenerationally^7,8^ , indicating that hot microclimates might function as reservoirs of heat adapted genotypes. Recent research even indicates that input of larvae from hotter reefs can boost bleaching tolerances on recipient reefs^9^, adding to the importance of locating these potential sources of heat adapted genotypes. Since these patches of heat adapted corals can probably occur on multiple scales, we need to consider both broad scale geographic and fine scale geomorphic factors as predictors of reef temperature regimes.  
  
```{r,echo=F}
setwd("C:/github/aeda_bos/raw-data/")
gbr_mmms<-read.csv("gbr_mmms.csv")
hawaii_mmms<-read.csv("hawaii_mmms.csv")
northaus_mmms <-read.csv("northaus_mmms.csv")
westaus_mmms <-read.csv("westaus_mmms.csv")

#Confirm that column names match
#table(colnames(hawaii_mmms) == colnames(gbr_mmms))
#table(colnames(northaus_mmms) == colnames(gbr_mmms))
#table(colnames(westaus_mmms) == colnames(gbr_mmms))

#Add "region" column
hawaii_mmms$REGION <- rep("HAWAII",nrow(hawaii_mmms))
gbr_mmms$REGION <- rep("GBR",nrow(gbr_mmms))
northaus_mmms$REGION <- rep("N_AUS",nrow(northaus_mmms))
westaus_mmms$REGION <- rep("W_AUS",nrow(westaus_mmms))

#Combine all regions into single dataframe
mmms_all <- bind_rows(gbr_mmms,northaus_mmms,westaus_mmms,hawaii_mmms)

#Convert latitude to absolute value
mmms_all$ABS_LAT <- abs(mmms_all$LATITUDE)

#Convert "region" to a factor
mmms_all$REGION <- as.factor(mmms_all$REGION)

#Change the order of the reef geomorphic class levels and convert it to a factor. Changing the order of the levels isn't really necessary, it just makes boxplots easier to interpret
rclass_order = c('Terrestrial Reef Flat','Shallow Lagoon','Deep Lagoon','Back Reef Slope','Inner Reef Flat','Outer Reef Flat','Reef Crest','Sheltered Reef Slope','Reef Slope','Plateau')
mmms_all$RCLASS <- factor(mmms_all$RCLASS,levels=rclass_order)

```
## Methods  
  
  This analysis relies on temperature data from submerged temperature loggers placed by the Australian Institute of Marine Sciences (AIMS) and the United States' National Oceanographic and Atmospheric Administration (NOAA).  
  NOAA's National Coral Reef Monitoring program deploys submerged SeaBird Electronics brand temperature loggers on coral reef throughout US and territorial waters^10^.  I downloaded the Coral Reef Monitoring program data through the NOAA National Centers for Environmental Information portal. The Hawaii data used in this project spans a timeframe from March 2005 to September 2019, though most of the individual points only have data for a subset of this time. Most loggers recorded data at 30 minute intervals.  
  I downloaded the AIMS data from the  AIMS Sea Water Temperature Observing System website. The data was recorded by a combination of Odyssey brand and Sensus Altra brand loggers, mostly recording temperature every five or ten minutes^11^. I used data points from January 2010 to present, though many of the individual logger points only include data for part of that timeframe.   
  For both the NOAA and the AIMS data, logger metadata included the depth for each submerged logger in meters, as well as latitude and longitude for each logger.  
  I downloaded geomorphic class data from the Allen Coral Atlas. All logger points used in this analysis fell in the Allen Coral Atlas regions of "Hawaiian Islands," "Great Barrier Reef and Torres Strait," "Timor and Arafura Seas," and "Western Australia." The Allen Coral Atlas relies on Planet Dove data, which in turn relies on imagery from Dove, Sentinel 2 and Landsat satellites. The geomorphic class algorithm also uses wave exposure data from the National Oceanic and Atmospheric Associate (NOAA) Wave Watch III global wave model. The Allen Coral Atlas creates maps of geomorphic classes using a combination of Random Forest classifiers and Object Based Analysis^12^.  
  All of the data preprocessing was run in Python. I used the package Geopandas to extract geomorphic class values to the logger points, and excluded points not associated with geomorphic classes. I also excluded days with fewer than 12 temperature logs, or gaps of more than six hours between temperature logs, then excluded months with fewer than 25 days worth of data. Finally, I downsampled all of the temperature logs at each point to even thirty minute spacing.  
  I used the preprocessed data to calculate maximum monthly mean, a statistic typically used to help assess coral bleaching likelihoods. Maximum monthly mean is a description of baseline summer temperature, and higher maximum monthly means are typically associated with higher bleaching thresholds, or lower probability of bleaching at a given temperature. I calculated maximum monthly mean by pooling all of the data for each month across at each logger and taking the mean. For example, if a given logger had data from January of 2012 to December of 2014, I calculated a January mean encompassing data from January 2012, January 2013, and January 2014, and so on for each month. Then, for each logger, I selected the month with the highest mean temperature. This is my maximum monthly mean temperature, abbreviated MMM.  
  I log transformed the depth variable before analysis to make it more normal.  
```{r, echo=FALSE}
ggplot(mmms_all, aes(DEPTH)) +
  geom_histogram(bins=15)+
  ggtitle("Logger depth") +
  labs(x = "Logger depth in meters")

mmms_all$LOG_DEPTH <- log1p(mmms_all$DEPTH)

ggplot(mmms_all, aes(LOG_DEPTH)) +
  geom_histogram(bins=12)+
  ggtitle("Logger depth") +
  labs(x = "Logger depth in log1p(meters)")

ggplot(mmms_all, aes(TEMP_C)) +
  geom_histogram(bins=15)+
  ggtitle("MMM temperatures") +
  labs(x = "Max monthly mean temp degrees C")
```

## Results  
  
  I visually examined the relationships between my independent variables and my dependent variables.  

The relationship between depth and MMM temperature lacks an obvious pattern.  
```{r message=TRUE, warning=FALSE, r,echo=FALSE}
ggplot(mmms_all,aes(x=LOG_DEPTH,y=TEMP_C))+
  geom_point(aes(color=as.factor(REGION)))+
  geom_smooth()+
  ggtitle("MMM temperature by logger depth")+
  labs(x = "Logger depth in log(meters)", y = "Max monthly mean temp degrees C")
```

Conversely, depth and MMM temperature have a clear, though nonlinear, relationship.  
```{r,echo=FALSE}
ggplot(mmms_all,aes(x=ABS_LAT,y=TEMP_C))+
  geom_point(aes(color=as.factor(REGION)))+
  geom_smooth()+
  ggtitle("MMM temperature by abs latitude")+
  labs(x = "absolute latitude", y = "Max monthly mean temp degrees C")
```

I ran some single variable linear models as a basic way of looking at the relationships between my variables.  
```{r,include=F}
lat_lm <-lm(TEMP_C ~ ABS_LAT, data=mmms_all)
#autoplot(lat_lm)
summary(lat_lm)

depth_lm <-lm(TEMP_C ~ LOG_DEPTH, data=mmms_all)
#autoplot(depth_lm)
summary(depth_lm)

rclass_lm <-lm(TEMP_C ~ RCLASS, data=mmms_all)
#autoplot(rclass_lm)
summary(rclass_lm)
```
```{r,echo=TRUE}
#Rsquared MMM temp ~ geomorphic class linear model
print(summary(rclass_lm)$r.squared)
#Rsquared MMM temp ~ depth linear model
print(summary(depth_lm)$r.squared)
#Rsquared MMM temp ~ absolute latitude linear model
print(summary(lat_lm)$r.squared)
```
Preliminarily, latitude looks important. Unfortunately, my data does not appear to meet all of the assumptions for linear models, and in particular does not demostrate a linear relationship between the dependent variable and each of its predictors.  

To avoid this issue, I also ran several nonparametric Generalized Additive Models (GAMs).I included absolute latitude as a smooth term, and depth and geomorphic class as parametric terms.  
```{r include=FALSE}
lat_gam <- gam(TEMP_C ~ s(ABS_LAT), data= mmms_all)
gam.check(lat_gam)
summary(lat_gam)

rclass_gam <-gam(TEMP_C ~ as.factor(RCLASS), data = mmms_all)
gam.check(rclass_gam)
summary(rclass_gam)

depth_gam <-gam(TEMP_C~LOG_DEPTH,data=mmms_all)
gam.check(depth_gam)
summary(depth_gam)

combo_gam <- gam(TEMP_C ~ s(ABS_LAT) + LOG_DEPTH + as.factor(RCLASS), data= mmms_all)
gam.check(combo_gam)
summary(combo_gam)

lat_rclass_gam <- gam(TEMP_C ~ s(ABS_LAT)+ as.factor(RCLASS), data= mmms_all)
gam.check(lat_rclass_gam)
summary(lat_rclass_gam)
```

I checked model assumptions using the gam.check function from the package "mgcv" and compared the models in term of both deviance explained and AIC.  
```{r,echo=T}
print(AIC(lat_gam,depth_gam,rclass_gam,combo_gam,lat_rclass_gam))
```

By AIC, the best model includes all three variables, though the depth variable never appears as significant.  
```{r,echo=TRUE}
print(summary(combo_gam))
```
The assumptions seem reasonably met.  
```{r,echo=TRUE}
gam.check(combo_gam)
```
I also ran a Random Forest model using the same data as the GAM models.  

```{r, echo=FALSE}
#Preprocess data for random forest
colnames(mmms_all)
mmms_keep <-mmms_all[,c(7,11,14,15)]

preprocesses <- preProcess(mmms_keep, method = 'medianImpute')
mmms_imp <- predict(preprocesses, mmms_keep)

folds <- createFolds(mmms_keep$TEMP_C, k = 5, returnTrain = TRUE)
trcntrl <- trainControl(index = folds, savePredictions = TRUE, search = 'random')

#Run random forest model
rf_gr <- expand.grid(mtry = c(1,2,3), splitrule = 'variance', min.node.size = c(5, 10, 20, 50))
mmm_rf <- train(TEMP_C ~ ., data = mmms_imp, method = 'ranger', tuneGrid = rf_gr, trControl = trcntrl, na.action = na.omit, importance = 'impurity', num.trees = 1000)

```

The Random Forest model ultimately produced a maximum R-squared value very similar to the best GAM model.  
```{r,echo=T}
print(mmm_rf$results$Rsquared %>% max)
```
Once again, absolute latitude appears as the most important variable.  
```{r,echo=T}
varImp(mmm_rf)
```
I also plotted the observed versus predicted variables for the Random Forest model.  
```{r,echo=F}
plotCV(mmm_rf)
```

## Discussion  
  
   In both GAM and random forest models, latitude was the most imporant predictor of maximum monthly mean temperature, far outweighing the effects of within-reef differences. Latitude alone accounted for a majority of the variation in MMM temperatures, though this might change once I incorporate data from other ocean basins. In general, temperature increased as absolute latitude decreased. However, the relationship between temperature and absolute latitude is notably nonlinear.  
	However, even though latitude was the most significant predictor, it was not the only significant predictor. I also found significant relationships between reef geomorphic class and MMM temperatures in boht the GAM and random forest models. This finding confirms that within-reef factors do have significant effects on temperature variation, even when accounting for latitude and depth. Likewise, the significance of geomorphic class in models including multiple geographic areas shows that morphologic features may have predictable effects across reefs.   
	Also interesting is the consistent lack of significance contribution from depth. However, this may arise in part from the limited range of depth values in the dataset: the Allen Coral Atlas geomorphic data only goes down to twenty meters, and I threw out the points below that depth that could not be associated with a geomorphic class, or even definitively confirmed as being of a reef. Likewise, I calculated maximum monthly mean temperatures using both diurnal and nocturnal temperature readings, as is standard. In the data cleaning process (not included in this document), I found that depth is much more associated with metrics of variability, like daily maximum minus daily minimum, than metrics of central tendency, like MMM.  
	Unfortunately, both GAM and Random Forest models have some notable limitations. First, their nonparametric nature makes it difficult to interpret relationships between variable quantitatively. Second, GAMs and Random Forest models can perform very poorly when predicting outside the original sample, especially if the new data to be used includes extreme values of the original predictors. Since an ultimate goal of this project is to predict temperature regimes in areas without in situ loggers, this constitutes a serious drawback.   
	One way to mitigate, or at least measure, the issue of predicting outside the sample is to divide the data to be fitted into a testing dataset and a training dataset. I intend to do this in the future, as I incorporate more geographic regions into the analysis. Right now, my limitation on incorporating NOAA data from the central Pacific and the Caribbean is that the Allen Coral Atlas is still in the process of creating geomorphic datasets for those regions; however, I hope to have them incorporated by the end of the summer at the latest. Furthermore, I am working on incorporating sea surface temperature data, an additional that I expect to radically change the model, possibly making latitude irrelevant. However, running latitude-based models was still a useful intellectual excerise in that it helps illucidate the relationship between geographic and geomorphic factors in governing temperature regimes on coral reefs.  
```{r}
```

## References  
  
1. Plaisance, L. et al. (2011) The Diversity of Coral Reefs: What Are We Missing? PLoS ONE 6, e25026  
2. van Hooidonk, R. et al. (2016) Local-scale projections of coral reef futures and implications of the Paris Agreement. Sci Rep 6, 39666  
3. McClanahan, T.R. et al. (2007) Effects of climate and seawater temperature variation on coral bleaching and mortality. Ecol. Monogr. 77, 503–525  
4. Craig, P. et al. (2001) High temperatures tolerated by a diverse assemblage of shallow-water corals in American Samoa. Coral Reefs 20, 185–189  
5. Howells, E.J. et al. (2016) Host adaptation and unexpected symbiont partners enable reef-building corals to tolerate extreme temperatures. Glob Change Biol 22, 2702–2714  
6. Schoepf, V. et al. (2015) Limits to the thermal tolerance of corals adapted to a highly fluctuating, naturally extreme temperature environment. Sci Rep 5, 17639  
7. Palumbi, S.R. et al. (2014) Mechanisms of reef coral resistance to future climate change. Science 344, 895–898  
8. Howells, E.J. et al. (2012) Coral thermal tolerance shaped by local adaptation of photosymbionts. Nature Clim Change 2, 116–120  
9. Kleypas, J.A. et al. (2016) Larval connectivity across temperature gradients and its potential effect on heat tolerance in coral populations. Glob. Change Biol. 22, 3539–3549  
10. Ecosystem Sciences Division, Pacific Islands Fisheries Science Center (2020). National Coral Reef Monitoring Program: Water temperature data from subsurface temperature recorders (STRs) deployed at coral reef sites in the Hawaiian Archipelago (NCEI Accession 0210383). NOAA National Centers for Environmental Information. Dataset.  
11. Australian Institute Of Marine Science Sea Temperature Observing System (SeaTemps Temperature Logger Program). . (2017) , Australian Institute of Marine Science  
12. Allen Coral Atlas (2020). Imagery, maps and monitoring of the world's tropical coral reefs. doi.org/10.5281/zenodo.3833242  